---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2025"
subtitle: "Assignment 2 - Due date 01/27/26"
author: "Zhaoxin Zhang"
output: pdf_document
geometry: margin=2.54cm
---

## Submission Instructions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github.   

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp26.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## R packages

R packages needed for this assignment:"forecast","tseries", and "dplyr". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=60), tidy=TRUE) 
library(forecast)
library(tseries)
library(dplyr)
library(ggplot2)
```


## Data set information  

Consider the data provided in the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx" on our **Data** folder. The data comes from the US Energy Information and Administration and corresponds to the December 2025 Monthly Energy Review. The spreadsheet is ready to be used. Refer to the file "M2_ImportingData_XLSX.Rmd" in our Lessons folder for instructions on how to read $.xlsx$ files.

```{r}
#Importing data set
library(readxl)
library(openxlsx)
getwd()
#Importing data set without change the original file using read.xlsx
energy_data1 <- read_excel(path="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",skip = 12, sheet="Monthly Data",col_names=FALSE) 

#Now let's extract the column names from row 11
read_col_names <- read_excel(path="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",skip = 10,n_max = 1, sheet="Monthly Data",col_names=FALSE) 

#Assign the column names to the data set
colnames(energy_data1) <- read_col_names

#Visualize the first rows of the data set
head(energy_data1)

```


## Question 1
You will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. Use the command head() to verify your data.

```{r}
energy_data_new <- energy_data1[, c(1, 4, 5, 6)]

head(energy_data_new)
```


## Question 2
Transform your data frame in a time series object and specify the starting point and frequency of the time series using the function ts().

```{r}
energy_data_tsa <- ts(energy_data_new[,2:4],start=c(1973,1),frequency=12)
energy_data_tsa
```


## Question 3
Compute mean and standard deviation for these three series.

```{r}
data_threeseries <- energy_data_new[, 2:4]
Mean <- colMeans(data_threeseries, na.rm = TRUE)
SD <- sapply(data_threeseries, sd, na.rm = TRUE)
Mean #Showing mean of the three series
SD #Showing standard deviation of the three series
```


## Question 4
Display and interpret the time series plot for each of these variables. Try to make your plot as informative as possible by writing titles, labels, etc. For each plot add a horizontal line at the mean of each series in a different color.

```{r}
#Mean
Mean <- colMeans(energy_data_tsa, na.rm = TRUE)
mean_tsa <- energy_data_tsa
mean_tsa[,] <- matrix(rep(Mean, each = nrow(energy_data_tsa)),
                      nrow = nrow(energy_data_tsa), byrow = FALSE)
autoplot(energy_data_tsa) +
  autolayer(mean_tsa, linetype = "dashed",alpha = 1.5, show.legend = FALSE) +
  ggtitle("Total Monthly Energy Production/Consumption") +
  xlab("Time") +
  ylab("Production/Consumption") +
  labs(color="Energy Type")
```
The Hydroelectric Power Consumption shows a releative constant and slow growth during the decades, while the biomass production and renewable energy production have similar trends. The biomass and renewable energy production have grown fast and fluctuated through years, which may due to the change of public policies and technology of renewable energy.

## Question 5
Compute the correlation between these three series. Are they significantly correlated? Explain your answer.

```{r}
correlation <- cor(energy_data_tsa, use = "complete.obs")
correlation
```
Total Biomass Production and Total Renewable Energy Production have a correlation of 0.96529851, showing that they are significantly correlated.
Total Biomass Energy Production and Hydroelectric Power Consumption have a correlation of -0.13473742, which is weak and not significant.
Total Renewable Energy Production and Hydroelectric Power Consumption have a correlation of -0.05842436, which also shows a weak and insignificant correlation.

## Question 6
Compute the autocorrelation function from lag 1 up to lag 40 for these three variables. What can you say about these plots? Do the three of them have the same behavior?

```{r}
Biomass_acf = Acf(energy_data_tsa[,1], lag.max = 40, plot = TRUE)
Biomass_acf
Renewable_acf = Acf(energy_data_tsa[,2], lag.max = 40, plot = TRUE)
Renewable_acf
Hydro_acf = Acf(energy_data_tsa[,3], lag.max = 40, plot = TRUE)
Hydro_acf
```
The total biomass production and renewable energy production trends have the similar behavior. Both total biomass production and renewable energy production show positive and significant autocorrelation from lag 1 up to lag 40, indicating that they have strong  high persistence and are correlated with their past values over time. 

The hydroelectric power consumption has a fluctuating trend: the autocorrelation moves up and down between positive and negative values, indicating a seasonal trend instead of a persistent trend.

## Question 7
Compute the partial autocorrelation function from lag 1 to lag 40 for these three variables. How these plots differ from the ones in Q6?

```{r}
Biomass_pacf = Pacf(energy_data_tsa[,1], lag.max = 40, plot = TRUE)
Biomass_pacf
Renewable_pacf = Pacf(energy_data_tsa[,2], lag.max = 40, plot = TRUE)
Renewable_pacf
Hydro_pacf = Pacf(energy_data_tsa[,3], lag.max = 40, plot = TRUE)
Hydro_pacf
```
All of the three variables show fluctuating trends between positive and negative values, and most of the lines show weak correlations. After removing the influence of all these intermediate variables, the trends of biomass production and renewable production have changed from what we saw in Q6. The biomass production and renewable energy production trends are very similar, indicating that they have similar seasonality. The hydroelectric consumption trend also show seasonality, while it has a more complex and fluctuating change through 12 months. 
